---
title: DSM analysis of fin whale sightings
author: David L Miller and Doug Sigourney

---

data provided by Doug Sigourney, NOAA NEFSC. based on data from the paper at https://peerj.com/articles/8226/

```{r data-and-packages}
library(mrds)
library(dsm)
load("findata.RData")
```




## Availability

```{r availability}
a_plane <- 0.37
a_ship <- 1

avail <- c(a_ship, a_plane)[fin_obs$survey]
```

```{r truncation}
w_ship <- 6
w_plane <- 0.9
```


## Model fitting

### Ship surveys - MRDS
```{r ship-mrds}
Ship_mrds <- ddf(dsmodel = ~mcds(key = "hr", formula = ~beaufort + SUBJ_WAVG),
                 mrmodel = ~glm(~distance),
                 data = ship_detections, method = "io",
                 meta.data = list(width = w_ship))
summary(Ship_mrds)
```


### Aerial surveys - MCDS

```{r plane-mcds}
Plane_ds <- ddf(dsmodel = ~mcds(key = "hr", formula = ~beaufort),
                data = plane_detections, meta.data = list(width = w_plane))
summary(Plane_ds)
```


### Density surface model

this is as in Sigourney et al, setting group sizes to be 1 then inflating with average group size later

```{r dsm-avg-group}
dsm_avg_group <- dsm(count ~ s(DIST125,bs="ts", k=5) +
                             s(DEPTH,bs="ts", k=5) +
                             s(DIST2SHORE,bs="ts", k=5) +
                             s(SST,bs="ts", k=5),
                     ddf.obj=list(Ship_mrds, Plane_ds),
                     family=tw(link="log"), method="REML",
                     segment.data=fin_segs, group=TRUE,
                     availability=avail,
                     observation.data=fin_obs)
```

```{r dsm-avg-group-checking}
summary(dsm_avg_group)
obs_exp(dsm_avg_group, "beaufort", c(0, 1, 2, 3, 4, 5))
```

```{r dsm-avg-group-var}
vp_b <- dsm_varprop(dsm_avg_group, predgrid)
vp_b
```

correct for group size:

Make predictions using predict function
mutiply each prediction by estimate of average group size
```{r dsm-avg-group-predict}
predict.fun <- predict(dsm_avg_group, newdata=predgrid,
                       off.set=predgrid$off.set)

sum(predict.fun*1.38)
```



same model but using observed rather than estimated group sizes

```{r dsm-obs-group}
dsm_obs_group <- dsm(count ~ s(DIST125,bs="ts", k=5) +
                             s(DEPTH,bs="ts", k=5) +
                             s(DIST2SHORE,bs="ts", k=5) +
                             s(SST,bs="ts", k=5),
                     ddf.obj=list(Ship_mrds, Plane_ds),
                     family=tw(link="log"), method="REML",
                     availability=avail,
                     segment.data=fin_segs,
                     observation.data=fin_obs)
```

```{r dsm-obs-group-checking}
summary(dsm_obs_group)
obs_exp(dsm_obs_group, "beaufort", c(0, 1, 2, 3, 4, 5))
```

better?
```{r dsm-obs-group-pred}
sum(predict(dsm_obs_group, newdata=predgrid, off.set=predgrid$off.set))
```

```{r dsm-obs-group-var}
# propagate uncertainty
vp <- dsm_varprop(dsm_obs_group, predgrid)
vp
```

